How to Avoid OOTA Without Really Trying

### 2024-09-02 paper/ootavacuous.tex
ootavacuous: Self-review part 1
Move compiler-related examples to the C++ Compilers section.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

	We focus on shared-memory programs and
	C++ implementations based on traditional compilers and
	computing hardware, including CPUs and GPGPUs.
	This context permits us to consider the detailed relations
	between source code and machine code required by the restricted
	nature of volatile atomic accesses.
	The OOTA problem and the related challenge of coming to grips
	with semantic dependency are much more tractable at the hardware
	level than at the source level, thanks to existing formal
	hardware models.
	We show that these models' constraints prevent OOTA cycles
	from occurring in undefined-behavior-free C++ programs running
	on compiler-based implementations, provided the cycles involve
	only volatile atomics.
	We also extend this work to nonvolatile atomics by defining
	``quasi-volatile'' behavior that we expect C++ implementations
	will adhere to if they perform single-thread analysis.
\end{abstract}
\maketitle{}

@@

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering.
An example of simple reordering is shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder P1's
lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior,
as indicated by the ``sometimes satisfied'' in the figure.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Some of these workers appear to have abandoned this effort, but as of
2024, Mark Batty is persisting with modular relaxed dependencies, which
may have the added ability of checking compiler optimizations.
@@@ cite web site.

Goldblatt looked at interactions between OOTA cycles and
undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
% Appendix~\ref{app:Aside on Undefined Behavior}
% analyzes his examples and notes ways to separate UB and OOTA-cycle
% concerns.
This document will concentrate on examples lacking UB.

@@

Section~\ref{sec:OOTA: rf versus rfe}
demonstrates advantages of formulating an OOTA cycle as a cycle in
sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
Section~\ref{sec:Properties of Semantic Dependencies}
then discusses properties of semantic dependences, showing that they
are functions of executions rather than strictly of source code,
among other things.

@@

In order to avoid these rfi links we will substitute rfe
for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe.
Any rfi links in a cycle can instead be interpreted as part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the interthread and intrathread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the intuitive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later in global time than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction in
Sections~\ref{sec:Global Optimization Can Destroy Dependencies}
and~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies} below.

@@

The C++ standard describes the execution of a program in terms of
``a parameterized nondeterministic abstract machine'' in 4.1.2p1
(\co{[intro.abstract]}).
This description specifies how the abstract machine carries out the
operations of a source program in great, but not complete, detail:
\begin{itemize}
\item	Some of the abstract machine's characteristics are
	implementation defined, including things like the number of
	bits in the various integer types
	or whether the \co{char} type is signed.
\item	Some aspects of an execution are unspecified or nondeterministic,
	including things like the order of evaluation of the operands
	of most binary operators or of the arguments in a function call.
	Implementations may choose from a set of allowed behaviors.
\item	Some actions are deemed to have undefined results; the standard
	says essentially nothing about programs that can give rise to
	undefined behavior.
\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
\item	Input and output are not described in any detail.
\end{itemize}
In addition to these points, the standard does not specify which store
an atomic load must read from, beyond requiring that the overall
pattern of loads and stores be consistent with the C++ memory model.
We assume that programs will not indulge in any computations that
could vary spontaneously from one execution to another,
such as basing a dependency on the time of day or a process ID.

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
running on implementations in
which \co{char} is a signed type,
but not those for which it is unsigned.

The same is true for the nondeterministic aspects of an execution.
Consider this example, with \co{i} initially zero:
\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}

@@

Under any particular implementation,
a single program can have many different abstract executions,
varying in their decisions about which store each load reads from
and thus the value obtained.
It's worth noting, however, that not all the possible abstract executions
of a program need be realizable by the machine-code executable file
produced by that implementation.
In fact, we will see that \emph{none} of the possible OOTA executions
allowed by the loose C++ abstract machine will ever be realized
by the executables produced by many compilers,

@@

\emph{Tagline: Compilers are constrained in optimizing atomics, especially
volatile atomics.}

@@

Section~\ref{sec:Users Influence the Behavior of Compilers}
shows that C++ compilers can be influenced by their users as well
as by the standard.
Section~\ref{sec:Global Optimization Can Destroy Dependencies}
presents an example showing that global optimizations can destroy
semantic dependencies, and then
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
analyzes examples showing that inventing atomic loads can destroy
semantic dependencies.
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
then presents required properties of volatile atomic operations, and
also defines properties of quasi-volatile atomic operations.

>>>>

	We focus on shared-memory programs and
	C++ implementations based on traditional compilers and
	computing hardware, including CPUs and GPGPUs.
	This context permits us to consider the detailed relations
	between source code and machine code required by the restricted
	nature of volatile atomic accesses.
	The OOTA problem and the related challenge of coming to grips
	with semantic dependency are much more tractable at the hardware
	level than at the source level, thanks to existing formal
	hardware models.
	We show that these models' constraints prevent OOTA cycles
	from occurring in undefined-behavior-free C++ programs running
	on compiler-based implementations, provided the cycles involve
	only volatile atomics.
	We also extend this work to nonvolatile atomics by defining
	``quasi-volatile'' behavior that we expect C++ implementations
	will adhere to if they confine themselves to single-thread analysis.
\end{abstract}
\maketitle{}

@@

It is important to distinguish true OOTA cycles from OOTA-like
behavior caused by simple reordering, such as that shown in
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ compiler and the CPU are within their rights to reorder P1's
lines~1 and~2, which can result in all of \co{x}, \co{y}, \co{r1},
and \co{r2} having the value 42 via the following steps.
First, P1's line~2 stores 42 to \co{x}.
Second, P0's line~1 loads 42 from \co{x} into \co{r1}.
Third, P0's line~2 stores \co{r1}, and thus 42, to \co{y}.
Fourth and finally, P1's line~1 loads 42 from \co{y} to \co{r2}.
Current C++ implementations can and do exhibit this reordering behavior,
as indicated by the ``sometimes satisfied'' in the figure.

@@

Other workers recommend various procedures to identify and avoid OOTA
cycles~\cite{Lahav:2017:RSC:3062341.3062352,Sinclair:2017:CAR:3079856.3080206,Lee:10.1145/3385412.3386010,MarkBatty2019ModularRelaxedDependenciesOOTA},
but none of these have been looked upon favorably by C++ implementers.
Many of these workers appear to have abandoned this effort, but as of
2024, Mark Batty is persisting with modular relaxed dependencies, which
may have the added ability of checking compiler optimizations.
@@@ cite web site.

Goldblatt looked at interactions between OOTA cycles and
undefined behavior (UB)~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
% Appendix~\ref{app:Aside on Undefined Behavior}
% analyzes his examples and notes ways to separate UB and OOTA-cycle
% concerns.
This paper focuses on examples lacking UB.

@@

Section~\ref{sec:OOTA: rf versus rfe}
demonstrates advantages of formulating an OOTA cycle as a cycle in
sdep $\cup$ rfe instead of the traditional sdep $\cup$ rf.
Section~\ref{sec:Properties of Semantic Dependencies}
then discusses properties of semantic dependences, showing that they
are not strict functions of source code.

@@

To avoid this issue, we will substitute rfe
for rf, defining an OOTA cycle---for now---as a cycle in sdep $\cup$ rfe.
Any rfi links in a cycle can instead be interpreted as part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the interthread and intrathread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the intuitive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later in global time than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
Later, we will see examples of this destruction.

@@

The C++ standard describes the execution of a program in terms of
``a parameterized nondeterministic abstract machine'' in 4.1.2p1
(\co{[intro.abstract]}).
This description specifies how the abstract machine carries out the
operations of a source program in great, but not complete, detail:
\begin{itemize}
\item	Some of the abstract machine's characteristics are
	implementation defined, including things like the number of
	bits in the various integer types
	or whether the \co{char} type is signed.
\item	Some aspects of an execution are unspecified or nondeterministic,
	including things like the order of evaluation of the operands
	of most binary operators or of the arguments in a function call.
	Implementations may choose from a set of allowed behaviors.
\item	Some actions are deemed to have undefined results; the standard
	says essentially nothing about programs that can give rise to
	undefined behavior.
\item	Asynchronous actions (i.e., signal handlers) are largely ignored.
\item	Input and output are not described in any detail.
\end{itemize}
In addition to these points, the standard does not specify which store
an atomic load must read from, beyond requiring that the overall
pattern of loads and stores be consistent with the C++ memory model.
We assume that programs will not indulge in any computations that
could vary spontaneously from one execution to another,
such as basing a dependency on the time of day or a process ID.
In short, the standard grants C++ implementations considerable freedom,
as detailed in
Section~\ref{sec:C++ Compilers}.

@@

Under any particular implementation,
a single program can have many different abstract executions,
varying in their decisions about which store each load reads from
and thus the value obtained.
It's worth noting, however, that not all the possible abstract executions
of a program need be realizable by the machine-code executable file
produced by that implementation.
In fact, we will see that \emph{none} of the possible OOTA executions
allowed by the loose C++ abstract machine will ever be realized
by the executables produced by many compilers.

@@

\emph{Tagline: Compilers have considerable freedom, but are constrained
in optimizing atomics, especially volatile atomics.}

@@

Section~\ref{sec:Architectures Influence the Behavior of Compilers}
show that machine architectures influence C++ compilers.
Section~\ref{sec:Optimizations Influence the Behavior of Compilers}
show, perhaps unsurprisingly, that optimizers influence C++ compilers.
Section~\ref{sec:Users Influence the Behavior of Compilers}
shows that compilers can also be influenced by their users.
Section~\ref{sec:Global Optimization Can Destroy Dependencies}
presents an example showing that global optimizations can destroy
semantic dependencies, and then
Section~\ref{sec:Inventing Atomic Loads Can Destroy Semantic Dependencies}
analyzes examples showing that inventing atomic loads can destroy
semantic dependencies.
Section~\ref{sec:Volatile and Quasi Volatile Accesses}
then presents required properties of volatile atomic operations, and
also defines properties of quasi-volatile atomic operations.

\subsection{Architectures Influence the Behavior of Compilers}
\label{sec:Architectures Influence the Behavior of Compilers}

The implementation-defined aspects of the standard permit architectures to
differ as to whether an abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
running on implementations in
which \co{char} is a signed type,
but not those for which it is unsigned.

\subsection{Optimizations Influence the Behavior of Compilers}
\label{sec:Optimizations Influence the Behavior of Compilers}

Compiler optimizers also exploit the nondeterministic aspects of an
execution.
Consider this example, with \co{i} initially zero:
\begin{quote}
\begin{verbatim}
int foo(int a, int b)
{
   return a / b;
}
